---
instaltitle: "Module 19"
author: "Laura Brubaker-Wittman"
date: "11/18/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# GIS and Spatial Analysis in R

Open following packages (make sure they are all installed as well!):

```{r}
library(adehabitatHR)
library(DT)
library(ggmap)
library(lubridate)
library(maptools)
library(readxl)
library(rgdal)
library(rgeos)
library(sf)
library(sp)
library(spatialEco)
library(spdplyr)
library(tidyverse)
library(zoom)
library(curl)
```

## Objectives:

    In this module, we use R to extend our use of the system to GIS analyses. This module will include working with GIS data to create maps, and to extract data from numerous layers of different data types for spatial analysis. This is a first step, and by no means comprehensive. Remember that R is constantly changing, and the ways and packages by which GIS analyses are being done are changing, too!
    
## Geographic Information Systems (GIS)

**Info on projections and coordinate systems**

## Challenge 1

Let’s explore this question using some actual data. First, load in the dataset “atelesranging.csv” and do some exploratory mapping and data analysis. This is only a little more complicated than how we’ve been doing it before given the spatial aspect of the data:

```{r}
library(curl)
f <- curl("https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/master/AN597_Fall19/atelesranging.csv")
d <- read.csv(f, header = TRUE, sep = ",")
head(d)
summary(d)
```

Let’s take a quick look at how these points map out:

```{r}
library(tidyverse)
ggplot(data = d, aes(x = UTMX, y = UTMY)) + geom_point() + theme_bw()
```

Ok, so we can see there’s some points… but without the projection and some other layers, they’re not much help. Let’s get some more information added to our ‘map’.

First, let’s make sure we’ve got the points in the proper projection. We can do that most simply using the {sf} package. Note that the crs command (which stands for ‘coordinate reference system’) in the st_as_sf call is where we designate the projection. In this case, 32718 is the EPSG code. To learn more about how to define the projection, see this helpful tutorial: *Me: Just went back to the module?*

In the meantime, let’s project!

```{r}
library(sf)
## Linking to GEOS 3.6.1, GDAL 2.1.3, PROJ 4.9.3
d <- st_as_sf(d, coords = c("UTMX", "UTMY"), crs = 32718)
```

Now, we’ll need to plot using the geom_sf() call, which will ensure that our ggplot is a map with the required coordinate system and projection (notice how our axes now contain the tranlsated latitude and longitude associated with our coordinate/projection):

```{r}
q <- ggplot() + geom_sf(data = d, color = "light pink", alpha = 0.5) + theme_bw() # Me: "alpha" is level of transperancy with 0 being completely transperant and 1 being compeltely opaque, so .5 is in between.
q
```

Now, to contextualize this data a bit more, we’ve also got some shapefiles associated with the study site. A shapefile is a simple, nontopological format for storing the location and attributes of a geographic feature, which can be represented by points, lines, or polygons (areas). Also associated with these values might be a dBASE table, which can store additional attributes associated with the shapefile features. In file format, shapefiles are actually several (5 or 6) files together. Most publicly available spatial layers (i.e., those available from governmental repositories) are available as shapefiles.

To load a shapefile into ***R***, we must load all associated files at once. Luckily, the package {rgdal} has a set command for doing so. Let’s load in a shapefile of the Tiputini River, which forms the southern boundary of the spider monkeys’ territory.

NOTE: To execute this code, you’ll need to download these files from the AN597_Fall19 folder on my GitHub page to a folder in your working directory called GISdata. *Me: Need to do this before can go on.*

